---
title: "Introduction to Jaatha"
author: "Paul Staab"
date: "jaatha `r packageVersion('jaatha')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{jaatha-introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE, include=FALSE}
set.seed(112233)
```


Jaatha is an estimation method that can use computer simulations to produce
maximum-likelihood estimates even when the likelihood function can not be
evaluated directly. It can be applied whenever it is feasible to conduct a lot 
of simulations, but works best when the data is at least approximately possion
distributed. 

Jaatha was orginially designed for demographic inference in evolutionary 
biology. Please refer to the vignette

```{r eval=FALSE}
vignette("jaatha-demographic-inference", package = jaatha)
```

if you are interested in this application.


The Algorithm
-------------

```{r}
data_obs <- c(2, 8, 0, 6, 1, 3, 2, 2, 0, 7)
```


Creating a Model
--------------
```{r}
library(jaatha)
sim_func <- function(x, y = NULL) rpois(10, x)
sim_func(c(1, 10))
sum_stats <- list(create_jaatha_stat("id", function(x, opts) x))

par_ranges <- matrix(c(0.1, 0.1, 10, 10), 2, 2)
rownames(par_ranges) <- c("x", "y")
colnames(par_ranges) <- c("min", "max")
par_ranges

jaatha_model <- create_jaatha_model(sim_func, par_ranges, sum_stats)
```


Preparing the Data
------------------
Use the `create_jaatha_data` function to prepare the observed data for being
used in Jaatha. The function expected the data to be in the format that is 
also returned by simulation function. In our example, `sim_func` returns a
numeric vector of length `10`, and `data_obs` already has the same format.
So we can import it

```{r}
jaatha_data <- create_jaatha_data(data_obs, jaatha_model)
```


Running Jaatha
--------------
```{r}
estimates <- jaatha(jaatha_model, jaatha_data, sim = 50, repetitions = 2)
estimates
```





