---
title: "Demographic Inference with Jaatha"
author: "Paul Staab"
date: "jaatha `r packageVersion('jaatha')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Jaatha for Demographic Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

When used for demographic inference, Jaatha supports using the package `coala` 
as simulation engine. If these packages do not suit your needs, it is of course 
also possible to use the normal interface described in the `Introduction` 
vignette.


Creating a Model
----------------
Jaatha automatically creates a simulation function, parameter ranges and 
summary statistics from a `coala` model. We can for example specify a simple 
isolation-with-migration model using `par_range`s to mark parameters we want
to estimate with Jaatha: 

```{r}
library(coala)
model <- coal_model(c(10, 15), 100) +
  feat_mutation(par_range("theta", 1, 10)) +
  feat_migration(par_range("m", 0, 3), symmetric = TRUE) +
  feat_pop_merge(par_range("t_split", 0.1, 2), 2, 1) + 
  feat_recombination(1) +
  sumstat_jsfs()
```

We can now just pass this `coala` model to the `create_jaatha_model` function
to convert it into a Jaatha model:

```{r}
library(jaatha)
jaatha_model <- create_jaatha_model(model)
```

This uses `coala` for the simulations, gets the parameter ranges specified with
`par_range` and uses summary statistics added to the model.
Coala supports a wide range of models. Please refer to its documentation for
more information.



Importing Data
--------------
You can use coala's `calc_sumstats_form_data` function to calculate the summary
statistic for genetic data. The output of this function can be directly passed
on to `create_jaatha_data`.



Running Jaatha
--------------
From here on, you can estimate parameters using the `jaatha` as described in the
introduction vignette.

